---
title: "Task1"
author: "Yuliia Lysaniuk"
date: "19/10/2025"
format:
  html:
    embed-resources: true
editor: source
---

Theme: My Cat Busya

### 1.1. Atomic Types

Create the following variables related to your chosen object:

-   **4 numeric** values (e.g., population, temperature, size, rating)\
-   **3 integer** values (e.g., number of moons, buildings, inhabitants)\
-   **2 logical** values (e.g., `has_water`, `is_inhabited`)\
-   **3 character** values (e.g., `name`, `region`, `category`)

**Tip:** Think of these variables as *attributes* of your object.\
Use `print()` or `cat()` to describe them in one sentence.

```{r}
# Numeric variables
weight <- 4.3            # in kilograms
age <- 3.5               # in years
sleep_hours <- 16.8      # average hours per day
jump_height <- 1.2       # in meters

# Integer variables
num_toys <- 7
num_meals <- 3
num_vet_visits <- 2

# Logical variables
is_lazy <- TRUE
likes_water <- FALSE

# Character variables
name <- "Busya"
breed <- "Scottish Fold"
favorite_food <- "chicken treats"
```

### 1.2. Vectors

Create three different vectors:

1.  **Numeric vector** combining all your numeric and integer variables.\
    *(Use `c()`.)*

2.  **Sequence vector** from any number between 10–50 to 100, Choose step between 2 and 10.\
    *(Use `seq()` or `:`.)*

3.  **Logical vector** of length 10 with random `TRUE`/`FALSE` values using `sample()`.

```{r}
# 1. Numeric vector (combine numeric + integer)
numeric_vector <- c(weight, age, sleep_hours, jump_height,
                    num_toys, num_meals, num_vet_visits)
print(numeric_vector)

# 2. Sequence vector (from 10 to 100, step 10)
sequence_vector <- seq(10, 100, by = 10)
print(sequence_vector)

# 3. Logical vector (10 random TRUE/FALSE)
logical_vector <- sample(c(TRUE, FALSE), 10, replace = TRUE)
print(logical_vector)
```

### 1.3. Matrix

Create a **3×3 matrix** with random numbers between **-50 and 50**\
(using `sample()` or `runif()`), and name its **rows and columns** according to your theme.

Then calculate row and column averages using `rowMeans()` and `colMeans()`.

```{r}
# Create 3x3 matrix with random numbers between -50 and 50
busya_matrix <- matrix(sample(-50:50, 9), nrow = 3, ncol = 3)

# Name rows and columns (related to Busya’s life)
rownames(busya_matrix) <- c("Morning", "Afternoon", "Evening")
colnames(busya_matrix) <- c("Playtime", "NapTime", "EnergyLevel")

print(busya_matrix)

# Row and column means
row_means <- rowMeans(busya_matrix)
col_means <- colMeans(busya_matrix)

cat("\nAverage by rows (daily periods):\n")
print(row_means)
cat("\nAverage by columns (activities):\n")
print(col_means)
```

### 1.4. List

Create a **list** that combines all your previous objects:\
atomic variables, vectors, and matrix. Add one extra element — a short **description** of your universe as a string.

```{r}
busya_list <- list(
  numerics = numeric_vector,
  sequence = sequence_vector,
  logicals = logical_vector,
  matrix = busya_matrix,
  description = "Busya is a playful Scottish Fold cat who loves naps, chicken treats, and observing birds from the window."
)
print(busya_list)
```

### 1.5. Factor

Create a **factor variable** representing categories within your universe, e.g.:

-   `"low"`, `"medium"`, `"high"`\
-   `"infant"`, `"child"`, `"adult"`\
-   `"poor"`, `"average"`, `"rich"`

Make sure you have **at least 5 observations** total.\
Then use `table()` to show how many objects fall into each category.

```{r}
# Factor of Busya’s mood levels during the day
mood_levels <- factor(
  c("happy", "sleepy", "playful", "sleepy", "happy", "grumpy"),
  levels = c("grumpy", "sleepy", "playful", "happy")
)
print(mood_levels)

# Frequency table
cat("\nBusya’s mood frequency:\n")
print(table(mood_levels))
```

### 1.6. Data Frame

Create a **data frame** related to your object that includes:

1.  **5–8 observations** (rows)\
2.  **At least 3 columns** of different data types\
    (e.g., name, numeric value, logical flag, category)

Then:

-   Rename the columns using `names()` or `colnames()`.\
-   Display its structure with `str()` and its summary with `summary()`.\
-   Add a new column derived from the others (for example, a ratio or condition).

```{r}
# Create a data frame about Busya’s daily routine
busya_day <- data.frame(
  Activity = c("Eating", "Sleeping", "Playing", "Grooming", "Watching birds"),
  Duration_hours = c(0.5, 16, 2, 1, 1.5),
  Enjoyment_level = c("high", "very high", "very high", "medium", "high"),
  Requires_owner = c(TRUE, FALSE, TRUE, FALSE, FALSE)
)

# Rename columns
colnames(busya_day) <- c("Activity", "Duration_h", "Enjoyment", "With_Owner")

# Display structure and summary
str(busya_day)
summary(busya_day)

# Add new column – is_long_activity (more than 1 hour)
busya_day$Is_Long <- busya_day$Duration_h > 1

cat("\nUpdated data frame (Busya's daily routine):\n")
print(busya_day)
```

## Task 2: Subsetting & Missing Data in Your Mini-Universe

### 2.1. Querying Vectors

Let's start with the vectors you created in **Task 1**.

2.1.1. From your **Sequence vector** (the one with all your numbers):

-   Select and print the 3rd element.
-   Select and print elements 2 through 5.
-   Select and print all values that are greater than 20 (or another number that makes sense for your data).

2.1.2. From your **Logical vector**:

-   Select and print only the TRUE values. (Hint: `your_vector[your_vector]`).
-   Use `which()` to get the indices (positions) of all the `TRUE` values.

```{r}
# 2.1.1 Sequence vector

cat("3rd element:", sequence_vector[3], "\n")
cat("Elements 2-5:", sequence_vector[2:5], "\n")
cat("Values > 20:", sequence_vector[sequence_vector > 20], "\n")

# 2.1.2 Logical vector

true_values <- logical_vector[logical_vector]
cat("TRUE values:", true_values, "\n")
true_indices <- which(logical_vector)
cat("Indices of TRUE values:", true_indices, "\n")

```

### 2.2. Querying Your "Universe" List

Use your main list from **Task 1**.

-   Access and print the **description** string using the `$` operator
-   Access and print your **3x3 matrix** using double brackets `[[ ]]`.
-   Access and print the **second element** from your **vector element** within the list. (This will be a "list-within-a-list" access).

```{r}
cat("Description:", busya_list$description, "\n")
cat("Matrix:\n")
print(busya_list[["matrix"]])
cat("Second element of numeric vector in list:", busya_list$numerics[2], "\n")

```

### 2.3. Investigating Your Data Frame

Use the **data frame** you created in **Task 1**.

-   Print the **first 3 rows**. (Use `head()`.)
-   Print the value from the **4th row, 1st column**.
-   Print only the column that contains your logical values (e.g., `is_habitable`) as a vector.

**Logical Subsetting**: Create a new data frame `df_subset` that contains only the rows that meet a specific logical condition.

-   Example: `df_subset <- your_df[your_df$population > 1000, ]`

`%in%` **Subsetting**: Create a new data frame `df_subset_2` that selects rows belonging to **two specific categories** from your factor or character column.

-   Example: `df_subset_2 <- your_df[your_df$region %in% c("North", "South"), ]`

```{r}
cat("First 3 rows:\n")
print(head(busya_day, 3))
cat("Value at [4,1]:", busya_day[4, 1], "\n")
cat("Logical column (With_Owner):\n")
print(busya_day$With_Owner)

# Logical subsetting

df_subset <- busya_day[busya_day$Duration_h > 1, ]
cat("Subset of long activities (>1h):\n")
print(df_subset)

# %in% subsetting

df_subset_2 <- busya_day[busya_day$Enjoyment %in% c("high", "very high"), ]
cat("Subset of activities with high enjoyment:\n")
print(df_subset_2)

```

### 2.4. Handling "Corrupted" Data (`NA`)

Create a copy of your original data frame (e.g., `df_corrupted <- your_df`).

**Introduce 3 `NA`s** into your df_corrupted at specific locations.

-   Example: `df_corrupted[2, 3] <- NA`

**Find `NAs`**:

-   Use `sum(is.na(df_corrupted)`) to find the **total number** of missing values.
-   Use `colSums(is.na(df_corrupted))` to see how many `NA`s are in **each column**.

**Calculate with `NA`s**:

-   Calculate the `mean()` of the numeric column where you added an `NA`.

**Filter `NA`s**:

-   Use `na.omit(df_corrupted)` to create a new, "clean" data frame that contains only the complete rows.

```{r}
df_corrupted <- busya_day
df_corrupted[2, "Duration_h"] <- NA
df_corrupted[5, "Enjoyment"] <- NA
df_corrupted[3, "With_Owner"] <- NA

cat("Corrupted Data Frame with NAs:\n")
print(df_corrupted)

total_na <- sum(is.na(df_corrupted))
cat("Total NAs:", total_na, "\n")
cat("NAs per column:\n")
print(colSums(is.na(df_corrupted)))

mean_duration <- mean(df_corrupted$Duration_h, na.rm = TRUE)
cat("Mean Duration_h (ignoring NA):", mean_duration, "\n")

df_clean <- na.omit(df_corrupted)
cat("Cleaned data frame (no NAs):\n")
print(df_clean)

```

### 2.5. Bonus Challenge: `which()`

Use `which.max()` or `which.min()` to find the **row number** (index) of the observation with the highest or lowest value in one of your numeric columns.

Use that index to print the row with that observation.

-   Example: `highest_pop_index <- which.max(your_df$population)`
-   Example: `your_df[highest_pop_index, ]`

```{r}
highest_duration_index <- which.max(busya_day$Duration_h)
cat("Row with highest Duration_h:\n")
print(busya_day[highest_duration_index, ])

```

## Task 3: Expanding Your Universe

**Dataset:** *World Cities — SimpleMaps*\
**Format:** CSV\
**Rows:** \~26,000\
**Columns:** 5\
**Data types:** character + numeric\

### 3.1. The Quest: Find Your Data

### 3.2. The Import: Load Your Data

```{r}
# Load libraries
library(dplyr)
library(readr)

# Import the dataset
world_cities <- read_csv("worldcities.csv")

# First inspection
glimpse(world_cities)
head(world_cities)
summary(world_cities)

```

### 3.3. The Interrogation: Explore with `dplyr`

\*\*5 Questions

**Question 1:** Which 10 cities have the largest population?

```{r}
world_cities %>% 
  select(city, country, population) %>% 
  arrange(desc(population)) %>% 
  head(10)

```

**Question 2:** What is the average population by country? (top 10 by population)

```{r}
world_cities %>% 
  group_by(country) %>% 
  summarise(avg_population = mean(population, na.rm = TRUE)) %>% 
  arrange(desc(avg_population)) %>% 
  head(10)

```

**Question 3:** How many cities exist for each capital type (primary, admin, minor, NA)?

```{r}
world_cities %>% 
  group_by(capital) %>% 
  summarise(count = n())

```

**Question 4:** Create a new variable population_millions and show the top 10 cities.

```{r}
world_cities %>% 
  mutate(population_millions = population / 1e6) %>% 
  select(city, country, population_millions) %>% 
  arrange(desc(population_millions)) %>% 
  head(10)

```

**Question 5:** Filter all cities with population \> 5 million and sort them from largest to smallest

```{r}
world_cities %>% 
  filter(population > 5e6) %>% 
  select(city, country, population) %>% 
  arrange(desc(population))

```

## Task 4. Building a "Toolkit" for Your Universe (Functions)

### 4.1. The "Greeting" Function (A Simple Start)

```{r}
 
welcome_to_my_universe <- function(your_name) {
  cat("--------------------------------------------\n")
  cat(" Welcome to the World Cities Analysis Toolkit\n")
  cat(" Report prepared by:", your_name, "\n")
  cat("--------------------------------------------\n")
}

welcome_to_my_universe("Yuliia")
```

### 4.2. The "Loading" Function

```{r}
library(readr)
library(dplyr)
library(ggplot2)

load_and_inspect <- function(file_path) {
  data <- read_csv(file_path)
  
  cat("--- Data loaded successfully from:", file_path, "---\n\n")
  
  cat("----- STRUCTURE -----\n")
  print(glimpse(data))
  
  cat("\n----- FIRST 6 ROWS -----\n")
  print(head(data))
  
  return(data)
}

# Load data from your folder
world_cities <- load_and_inspect("C:/Users/ylysa/Desktop/pocpodpora/worldcities.csv")

```

### 4.3. The "Query" Function

```{r}
filter_cities_by_population <- function(data, pop_threshold) {
  result <- data %>% 
    filter(population > pop_threshold) %>% 
    select(city, country, population) %>%
    arrange(desc(population))
  
  return(result)
}

big_cities <- filter_cities_by_population(world_cities, 5000000)
print(big_cities)


```

### 4.4. The "Reporting" Function

```{r}
generate_report <- function(data, country_name) {
  
  country_data <- data %>% 
    filter(country == country_name)
  
  if (nrow(country_data) == 0) {
    cat("No cities found for:", country_name, "\n")
    return(NULL)
  }
  
  city_count <- nrow(country_data)
  max_pop <- max(country_data$population, na.rm = TRUE)
  avg_pop <- mean(country_data$population, na.rm = TRUE)
  min_pop <- min(country_data$population, na.rm = TRUE)
  
  cat("----------- REPORT FOR:", country_name, "-----------\n")
  cat("Number of cities:", city_count, "\n")
  cat("Maximum population:", max_pop, "\n")
  cat("Average population:", round(avg_pop, 2), "\n")
  cat("Minimum population:", min_pop, "\n")
  cat("-----------------------------------------------\n\n")
}

generate_report(world_cities, "Ukraine")
generate_report(world_cities, "Slovakia")
```


## Task 5. Visualising Your Universe (Base R & ggplot2)
### 5.1. The Quick Sketch (Base R Histogram)

```{r}
library(dplyr)
library(readr)
library(ggplot2)

# Load again (optional)
world_cities <- read_csv("C:/Users/ylysa/Desktop/pocpodpora/worldcities.csv")

```

Sometimes you need a quick look at the distribution of your data without loading heavy packages.

  1. Choose one **numeric variable** from your dataset (e.g., `price`, `temperature`, `speed`).

  2. Use the `hist()` function from base R.

  3. Customise the `main` title, `xlab`, and the bar `col` (colour).
  
```{r}
hist(world_cities$population,
     main = "Distribution of City Populations",
     xlab = "Population",
     col = "lightblue",
     border = "white")

```
  
  ### 5.2. The Relationship Map (ggplot2 Scatter Plot)

Now, let's look for patterns using the industry-standard package, **ggplot2**.

  1. Choose **two numeric variables** to see if they are correlated.

  2. Create a scatter plot using `geom_point()`.
  
  3. Add `geom_smooth()` to add a relationship between variables to the plot.

  4. Add `labs()` to rename your axes and title.
  
```{r}
ggplot(world_cities, aes(x = lat, y = population)) +
  geom_point(alpha = 0.6) +
  geom_smooth() +
  labs(title = "Relationship Between Latitude and Population",
       x = "Latitude",
       y = "Population") +
  theme_minimal()
```

### 5.3. The Comparison (ggplot2 Boxplot)

3. The Comparison (ggplot2 Boxplot)

Boxplots are excellent for comparing groups. Let's see how different categories in your universe perform.

  1. Choose a **categorical (or logical) variable** for the X-axis (e.g., city, planet_type, species).

  2. Choose a **numeric variable** for the Y-axis (e.g., price, gravity, lifespan).

  3. Create a plot using `geom_boxplot()`.

  4. **Bonus:** Use `fill` to colour the boxes by the same category.
  
```{r}
ggplot(world_cities, aes(x = capital, y = population, fill = capital)) +
  geom_boxplot() +
  labs(title = "Comparison of Population by Capital Status",
       x = "Capital Status",
       y = "Population") +
  theme_light() +
  theme(legend.position = "none")

```

### 5.4. The "Deep Dive" (Faceting or Custom)

The most powerful feature of `ggplot2` is faceting—creating small multiples of the same plot for different groups.

  1. Create a plot (histogram, bar plot, or scatter plot).

  2. Use `facet_wrap(~ your_category_variable)` to split the plot into a separate panel for each group in your data.

  3. This is incredibly useful for comparing distinct groups (e.g., separate charts for each City or Year).

```{r}
world_cities_small <- world_cities |> 
  filter(country %in% c("Slovakia", "Ukraine", "Poland", "Germany", "France"))

ggplot(world_cities_small, aes(x = population)) +
  geom_histogram(bins = 15, fill = "coral", color = "black") +
  facet_wrap(~ country) +
  labs(title = "Population Distribution Across Selected Countries",
       x = "Population",
       y = "Count")

```

### 5.5. Final Step: Exporting (Saving Your Work)

A plot is useless if you can't put it in a report.

  1. Choose your favourite plot from the exercises above.

  2. Run the code to display it in R.

  3. Immediately run the following command to save it to your project folder:
  
```{r}
final_plot <- ggplot(world_cities, aes(x = lat, y = population)) +
  geom_point()

ggsave("C:/Users/ylysa/Desktop/pocpodpora/my_universe_plot.png",
       plot = final_plot,
       width = 8,
       height = 6)
```

